Univalent Universes and Completness of Reversible Programs: \\ Featherweight HoTT

### 2017-03-07 upi.tex
cleaned up definitions

\title{Univalent Universes and Completness of Reversible Programs: \\
  Featherweight HoTT}

@@

[Following the HoTT book and the slides on ``A Characterization of Univalent
Fibrations'' by Dan Christensen.]

@@

\paragraph*{Transport.} Given a type family $P : A \to \mathcal{U}$, a path
$p : x \equiv y$ in $A$ induces a function $p_* : P(x) \to P(y)$ which
transports points in $P(x)$ to points in $P(y)$.

\paragraph*{Fibration.} We think of a type family $P : A \to \mathcal{U}$ as a
\emph{fibration} with base space $A$, with $P(x)$ being the fiber over $x$, and
with $\Sigma_{(x:A)} P(x)$ being the \emph{total space} of the fibration, and
with first projection $\Sigma_{(x:A)} P(x) \to A$. In the figure below, $p_*(u)$
is the transport of $u$ along $p$ and both pairs $(x,u)$ and $(y,p_*(y))$ are in
the total space. A critical property of fibrations is that a path
$p : x \equiv y$ in the base space $A$ and a point $u$ in the fiber over $x$
induce a function that \emph{lifts} the path to the total space: the lifted path
starts at $(x,u)$ and ends at $(y,p_*(u))$. A function $f : \Pi_{(x:A)} P(x)$ is
sometimes called a \emph{section} of the fibration $P$.

@@

It is worth repeating Remark 2.7.1 from the book. In the figure above, we could
have $(x,u) \equiv (x,v)$ without having $u \equiv v$. The only thing we can
conclude is that there is a (possibly non-trivial) path path $q : x \equiv x$
such that $q_*(u) \equiv v$.

\paragraph*{Univalent Type Family.}
Given standard constructions in the book, we have a function:
\[\begin{array}{rcl}
\textit {transport-equiv} &:& \Pi_{(P : A \to \mathcal{U})}~ \Pi_{(x,y:A)}~
    x \equiv y  \to P x \simeq P y \\
\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoeqv}(\mathit{ap}_{P}(p))
\end{array}\]
with the standard definitions for
$\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y))$ and
$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$. We say a type family
$P : A \to \mathcal{U}$ is \emph{univalent} if for all $x,y:A$
the function $\textit{transport-equiv}~P~x~y$ is an equivalence. Formally:
\[\begin{array}{l}
\textit{is-univalent-fibration} : (A \to \mathcal{U}) \to \mathcal{U} \\
\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{is-equiv}~(\textit{transport-equiv}~P~x~y)
\end{array}\]

\noindent\paragraph*{Examples}
\begin{itemize}
\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $\Pi_{(x,y:\mathcal{U})}~\textit{is-equiv}~\mathit {idtoeqv}$ which is the
  usual definition of univalence.

\newpage

\item Take $A = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
  statement $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
\item Take $A = \ot$ and $P = \lambda \_. \zt$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to
  $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
\item Take $A = \ot$. For no $P$ other than the above two instances will
  we have a univalent fibration.
\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \_. \ot$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\ot\simeq\ot)$ which is false.
\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\zt\simeq\ot)$ which is true. That is the only choice of $P$
  that works (modulo symmetry).
\item For any $n > 2$, taking $A$ as sum of $n$ copies of $\ot$ can never
  give a univalent fibration.
\end{itemize}

\begin{theorem}
  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
  if there exists an $F$ such that $X \simeq \Sigma_{Y : \mathcal{U}} \| Y \equiv F
  \|$. We will refer to this type as $\{F\}$.
\end{theorem}

\noindent In particular $\{\bt\}$, i.e.,
$\Sigma_{Y : \mathcal{U}} \| Y \equiv \bt \|$ is the base of a univalent
fibration. In some sense this is the simplest univalent universe, our
featherweight HoTT that we aim to characterize as a reversible programming
language based on type isomorphisms.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Some Simple Cases}

Pick $F = \ot$. We have $\{\ot\} \simeq \ot$. Another
simple but less trivial case is $F = S^1$. In that case we also have
$\{S^1\} \simeq S^1$.p

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Bool Case}

To get started let's look at the case of $\bt$ instead of the entire set of finite types defined using $0$, $1$, $+$, and $*$.

@@

The restriction of $\Pi$ is the following:

@@

%%%%%
\subsection{$\mathcal{U}$}

We assume an ambient univalent universe defined like in the HoTT book and including dependent pairs, a unit type, coproducts, identity types, and constants $\textsc{funext}$ and $\textsc{univalence}$. Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \bt \|$ where:

\item $\Sigma$ is the type former for dependent pairs: it comes equipped with an introduction rule, an induction principle, and a computational rule.

\item $\bt$ is the type of booleans which comes with the two points, an induction principle, and computational rules.

\item $\equiv$ is the identity type former which comes with $\reflp$, an induction principle $J$, and a computation rule.

\item $\| A \|$ is the propositional truncation of $A$ defined as follows: for
  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have
  $\gluep : x \equiv y$.

@@

Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where
$\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points,
combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators
$\alpha : c_1 \isotwo c_2$ as 2-paths. The proof is in a meta-logic which is
itself univalent.

>>>>

\title[Featherweight HoTT]{Univalent Universes and Completness of Reversible Programs: \\
  Featherweight HoTT}

@@

Following the HoTT book and the slides on ``A Characterization of Univalent
Fibrations'' by Dan Christensen.

%%%%%
\subsection{Definitions}

@@

\paragraph*{Functors (Lemma 2.2.1 in the book)} Suppose $f : A \to B$ is a
function. Then for any $x,y:A$ there is an operation
$\mathit{ap}_f : (x \equiv y) \to (f(x) \equiv f(y))$.

\paragraph*{Equivalences (Chapter 4 in the book)} For any two types $A$ and $B$
we can form the type $A \simeq B = \Sigma_{(f : A \to B)} \mathit{isequiv}~f$ of
equivalences. Each equivalence consists of a function $f : A \to B$ together
with evidence that this function is an equivalence.

\paragraph*{Paths to Equivalences (Lemma 2.10.1 in the book)} Given two types
$A$ and $B$, there is a function
$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$ mapping paths to
equivalences.

\paragraph*{Transport (Lemma 2.3.1 in the book)} Given a type family
$P : A \to \mathcal{U}$, a path $p : x \equiv y$ in $A$ induces a function
$p_* : P(x) \to P(y)$ which transports points in $P(x)$ to points in $P(y)$.

\paragraph*{Fibration (Sections 2.3 and 2.7 in the book)} We think of a type
family $P : A \to \mathcal{U}$ as a \emph{fibration} with base space $A$, with
$P(x)$ being the fiber over $x$, and with $\Sigma_{(x:A)} P(x)$ being the
\emph{total space} of the fibration, and with first projection
$\Sigma_{(x:A)} P(x) \to A$. In the figure below, $p_*(u)$ is the transport of
$u$ along $p$ and both pairs $(x,u)$ and $(y,p_*(y))$ are in the total space. A
critical property of fibrations is that a path $p : x \equiv y$ in the base
space $A$ and a point $u$ in the fiber over $x$ induce a function that
\emph{lifts} the path to the total space: the lifted path starts at $(x,u)$ and
ends at $(y,p_*(u))$. A function $f : \Pi_{(x:A)} P(x)$ is sometimes called a
\emph{section} of the fibration $P$.

@@

\noindent It is worth repeating Remark 2.7.1 from the book. In the figure above,
we could have $(x,u) \equiv (x,v)$ without necessarily having $u \equiv v$. The
only thing we can conclude regarding $u$ and $v$ is that there is a (possibly
non-trivial) path $q : x \equiv x$ such that $q_*(u) \equiv v$.

\paragraph*{Propositional Truncation.} Given a type $A : \mathcal{U}$, the type
$\|A\|$ is the propositional truncation of $A$ defined as follows: for
  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have
  $\gluep : x \equiv y$.

\paragraph*{Univalent Type Family.}
For each fibration, we have a function:
\[\begin{array}{rcl}
\textit {transport-equiv} &:& \Pi_{(P : A \to \mathcal{U})}~ \Pi_{(x,y:A)}~
    x \equiv y  \to P(x) \simeq P(y) \\
\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoeqv}(\mathit{ap}_{P}(p))
\end{array}\]
We say a type family $P : A \to \mathcal{U}$ is \emph{univalent} if for all
$x,y:A$ the function $\textit{transport-equiv}~P~x~y$ is an
equivalence. Formally:
\[\begin{array}{l}
\textit{is-univalent-fibration} : (A \to \mathcal{U}) \to \mathcal{U} \\
\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{isequiv}~(\textit{transport-equiv}~P~x~y)
\end{array}\]

%%%%%
\subsection{Examples}

We will choose particular base spaces $A$ and fibrations $P : A \to \mathcal{U}$
below and reason about whether they are univalent or not. For each fibration,
the question of whether it is univalent reduces to whether, for all $x,y:A$, the
function $\textit{transport-equiv}~P~x~y$ is an equivalence. In other words, the
question we are asking is whether we have an equivalence
$(x \equiv y) \simeq (P(x) \simeq P(y))$ witnessed by the function
$\textit{transport-equiv}~P~x~y$.

\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. By definition,
  this fibration is univalent if
  $\textit{transport-equiv}~\mathit{Id}_{\mathcal{U}}~x~y = \mathit{idtoeqv}$ of
    type $x \equiv y \to x \simeq y$ is an equivalence. As Axiom 2.10.3 in the
    book states this is exactly the statement of the conventional univalence
    axiom for universe $\mathcal{U}$.

\item Take $A = \ot$ with element $\star$ and $P = \lambda \_. \ot$. This
    fibration $P$ is univalent if we have an equivalence
    $(\star\equiv\star) \simeq (\ot\simeq\ot)$ witnessed by
    $\textit{transport-equiv}~P~x~y$. The two types are clearly equivalent in
    this case.

\item Take $A = \ot$ and $P = \lambda \_. \zt$. This fibration is univalent if
    we have an equivalence $(\star\equiv\star)\simeq(\zt\simeq\zt)$ witnessed by
    $\textit{transport-equiv}~P~x~y$. The two types are also clearly equivalent
    in this case.

\item Take $A = \ot$. For no $P$ other than the above two instances will we
    have a univalent fibration.

\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \_. \ot$. The statement
  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
  (\ot\simeq\ot)$ which is false.

\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$. This fibration is univalent if
  we have an equivalence $(\fc \equiv \tc) \simeq (\zt\simeq\ot)$ which is
  true. That is the only choice of $P$ that works (modulo symmetry).

\item For any $n > 2$, taking $A$ as sum of $n$ copies of $\ot$ can never
  give a univalent fibration.

\item For any type $F$ we define
  $\{F\} = \Sigma_{A : \mathcal{U}} \| F \simeq A \|$. By a proposition of
  Christensen, any fibration with base space $\{F\}$, i.e., any fibration
  $\{F\} \to \mathcal{U}$, is univalent. In particular, we can take the base
  space to be:
  \begin{itemize}
  \item $\{\ot\}$. The type $\{\ot\}$ is actually the same as $\ot$ and this was
    covered by previous examples.
  \item $\{S^1\}$. The type $\{S^1\}$ is the same as $S^1$. This is perhaps the
    simplest base space to choose but it does not naturally correspond to a
    conventional programming language.
  \item $\{\bt\}$. In some sense this is the simplest
    programming-language--relevant universe --- our featherweight HoTT. We aim
    to characterize this base space as a reversible programming language based
    on type isomorphisms.
  \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analyzing $\{\bt\}$ and its Connection to $\Pi$}

%%%%%
\subsection{$\{\bt\}$}

Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$.

@@

The restriction of $\Pi$ to the case of just booleans is the following:

@@

@@

Now we claim that $\{\bt\}$ is equivalent to $\mathcal{U}_\Pi$ which is the
universe which includes types $\tau$ as a points, combinators
$c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators
$\alpha : c_1 \isotwo c_2$ as 2-paths. The proof is in a meta-logic which is
itself univalent.
